---
title: "TOPIC 1"
author: "Flavie Bertrand and Johanna Rakotovao"
date: "2025-11-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(knitr)
library(kableExtra)

```


# Data managing

## Import data 

```{r}
getwd()
citic300 <- read.csv("C:/Users/flavi/OneDrive/Documents/ENSIIE/Eco/S&P_CITIC300 Historical Data.csv")
sp500 <- read.csv("C:/Users/flavi/OneDrive/Documents/ENSIIE/Eco/S&P 500 Historical Data.csv")
head(sp500)
head(citic300)
dim(sp500)
dim(citic300)
```

## Merge data

Modify names of the variables :
```{r}
# For the US
sp500new<-sp500
sp500new$price_us<-sp500new$Price
sp500new$open_us<-sp500new$Open
sp500new$high_us<-sp500new$High
sp500new$low_us<-sp500new$Low
sp500new$vol_us<-sp500new$Vol.
sp500new$change_us<-sp500new$Change..
sp500new<-subset(sp500new, select = -c(Price, Open, High, Low, Vol., Change..))
head(sp500new)

# For China
citic300new<-citic300
citic300new$price_china<-citic300new$Price
citic300new$open_china<-citic300new$Open
citic300new$high_china<-citic300new$High
citic300new$low_china<-citic300new$Low
citic300new$vol_china<-citic300new$Vol.
citic300new$change_china<-citic300new$Change..
citic300new<-subset(citic300new, select = -c(Price, Open, High, Low, Vol., Change..))
head(citic300new)
```

Merge the two dataset :
```{r}
sp500new <- sp500new[order(sp500new$Date), ]
citic300new <- citic300new[order(citic300new$Date), ]
merged <- merge(sp500new, citic300new, by = "Date")
merged <- merged[order(merged$Date), ]
head(merged)
str(merged)
dim(merged)
```

## Cleaning data

```{r}
# Good format for date
merged$Date <- mdy(merged$Date)
merged <- merged[order(merged$Date), ]
head(merged)

# Cleaning price : replace the comma with nothing
merged$price_us = as.numeric(gsub(",", "", merged$price_us))
merged$price_china = as.numeric(gsub(",", "", merged$price_china))

# Cleaning of the change of %
merged$change_us = as.numeric(gsub("%", "", merged$change_us)) / 100
merged$change_china = as.numeric(gsub("%", "", merged$change_china)) / 100

head(merged)
str(merged)
dim(merged)
rownames(merged) <- NULL
head(merged)
dim(merged)

data_final<-subset(merged, select = c(Date, price_us, price_china, change_us, change_china))
view(data_final)
head(data_final)
```


# Computation of the volatility

## Compute the log return

```{r}
data_final_log <- data_final %>%
  arrange(Date) %>%
  mutate(
    log_ret_us = c(NA, diff(log(price_us))),
    log_ret_china = c(NA, diff(log(price_china)))
  ) %>%
  filter(!is.na(log_ret_us) & !is.na(log_ret_china))

head(data_final_log)
```


## Separation of the periods

```{r}
pre_pandemic <- subset(data_final_log, Date >= "2015-01-01" & Date < "2020-01-01")
post_pandemic <- subset(data_final_log, Date >= "2020-01-01" & Date <= "2024-12-31")

# Convert Date and filter by period
data_final_log$Date <- as.Date(data_final_log$Date, format="%d/%m/%Y")

# Split pre- and post-COVID periods
data_pre  <- subset(data_final_log, Date >= "2015-01-01" & Date < "2020-01-01")
data_post <- subset(data_final_log, Date >= "2020-01-01" & Date <= "2024-12-31")

# Extract log returns
r_us_pre  <- data_pre$log_ret_us
r_us_post <- data_post$log_ret_us
r_ch_pre  <- data_pre$log_ret_china
r_ch_post <- data_post$log_ret_china
```




## Use of GARCH(1,1) model to compute volatility


In financial data, volatility is not stable over time, and that is why we need to use a GARCH(1,1) model to compute the volatility. The volatility is time-varying.

In a GARCH(1,1) model, we have this :

- $\epsilon_t = \sigma_t Z_t$

- $\sigma_t^2 = \omega + \alpha \epsilon_{t-1}^2 + \beta \sigma_{t-1}^2$

Where :

- $\sigma_t^2$ is the conditional variance, so the volatility at time t give, the past

- $\alpha$ : impact of a return on the forthcoming volatility

- $\beta$ : memory effect on the volatility

```{r}
# Function to compute conditional variance and log-likelihood
garch11_loglik <- function(params, returns) {
  omega <- params[1]
  alpha <- params[2]
  beta  <- params[3]
  
  n <- length(returns)
  sigma2 <- rep(var(returns), n)
  eps2 <- returns^2
  
  # Recursively compute conditional variance
  for (t in 2:n) {
    sigma2[t] <- omega + alpha * eps2[t-1] + beta * sigma2[t-1]
  }
  
  # Compute conditional log-likelihood
  loglik <- -0.5 * sum(log(2 * pi) + log(sigma2) + eps2 / sigma2)
  return(-loglik)  # negative for minimization
}

# Estimate manually
init <- c(omega = 0.0001, alpha = 0.05, beta = 0.9)
res_us_pre <- optim(init, garch11_loglik, returns = r_us_pre, method = "L-BFGS-B",
                    lower = c(1e-6, 0, 0), upper = c(1, 1, 1))
res_us_post <- optim(init, garch11_loglik, returns = r_us_post, method = "L-BFGS-B",
                     lower = c(1e-6, 0, 0), upper = c(1, 1, 1))

res_ch_pre <- optim(init, garch11_loglik, returns = r_ch_pre, method = "L-BFGS-B",
                    lower = c(1e-6, 0, 0), upper = c(1, 1, 1))
res_ch_post <- optim(init, garch11_loglik, returns = r_ch_post, method = "L-BFGS-B",
                     lower = c(1e-6, 0, 0), upper = c(1, 1, 1))


# Compute variance 

get_sigma2 <- function(params, returns) {
  omega <- params[1]; alpha <- params[2]; beta <- params[3]
  n <- length(returns)
  sigma2 <- rep(var(returns), n)
  for (t in 2:n) sigma2[t] <- omega + alpha * returns[t-1]^2 + beta * sigma2[t-1]
  return(sigma2)
}

sigma_us_pre  <- get_sigma2(res_us_pre$par, r_us_pre)
sigma_us_post <- get_sigma2(res_us_post$par, r_us_post)
sigma_ch_pre  <- get_sigma2(res_ch_pre$par, r_ch_pre)
sigma_ch_post <- get_sigma2(res_ch_post$par, r_ch_post)

```

# Block Bootstrap test function

We cannot just use use a ttest to estimate the volatility. The problem is that the $\sigma_t^2$ values come from a time series, so they are autocorrelated, and a ttest has an independent hypothesis about the observations.

To know if volatility changed between to periods, we will test : 

- $H_0 : E[\hat\sigma_{1,t}^2] = E[\hat\sigma_{2,t}^2]$,

- $H_1 : E[\hat\sigma_{1,t}^2] \neq E[\hat\sigma_{2,t}^2]$.

But the variances are autocorrelated, since by definition they depend on past observations.

We tried to find another way "by hand" to do it.
In that way, we tried to simulate the null hypothesis empirically by using a bootstrap.

In bootstrap, we resample blocks of consecutive days and then compare the average volatility between groups a lot of time.

We will have as a consequence an empirical result of the differences of volatility and a p-value.

Here is how we proceeded : 

1. We divided each volatility series into blocks of a certain length.

2. We resample randomly these blocks with replacement to form new bootstrap samples.

3. For each bootstrap replication $b=1,…,B$
b=1,…,B, we compute the difference in mean volatility.

4. The empirical distribution of the difference in mean volatility approximates the null distribution of the test statistic with dependence.
The p-value is computed : 
$$
p = \frac{1}{B} \sum_{b=1}^{B} \mathbb{1}\left( |\Delta^{*(b)}| \geq |\Delta_{\text{obs}}| \right)
$$

where 
$$ 
\Delta_{\text{obs}} = \bar{\sigma}_{1}^{2} - \bar{\sigma}_{2}^{2} 
$$ 
is the observed difference in mean volatility, 
$$
\Delta^{*(b)} = \bar{\tilde{\sigma}}_{1}^{2*(b)} - \bar{\tilde{\sigma}}_{2}^{2*(b)}
$$ 
is the difference in mean volatility and $\tilde{\sigma}_{1,t}^{*(b)}$ 
and $\tilde{\sigma}_{2,t}^{*(b)}$ are the new bootstrap samples.


If the p-value is small ($p<0.5$), we reject $H_0$ and conclude that the average volatility is significantly different between two periods.

```{r}
block_bootstrap_test <- function(x, y, block_size = 20, nboot = 1000) {
  n_x <- length(x); n_y <- length(y)
  mean_diff <- mean(x) - mean(y)
  diffs <- numeric(nboot)
  
  for (i in 1:nboot) {
    # resample contiguous blocks from each series
    start_x <- sample(1:(n_x - block_size + 1), 1)
    start_y <- sample(1:(n_y - block_size + 1), 1)
    boot_x <- x[start_x:(start_x + block_size - 1)]
    boot_y <- y[start_y:(start_y + block_size - 1)]
    diffs[i] <- mean(boot_x) - mean(boot_y)
  }
  
  pval <- mean(abs(diffs) >= abs(mean_diff))
  return(list(mean_diff = mean_diff, pval = pval, diffs = diffs))
}

# Run tests
boot_us_prepost   <- block_bootstrap_test(sigma_us_pre,  sigma_us_post)
boot_ch_prepost   <- block_bootstrap_test(sigma_ch_pre,  sigma_ch_post)
boot_us_ch_pre    <- block_bootstrap_test(sigma_us_pre,  sigma_ch_pre)
boot_us_ch_post   <- block_bootstrap_test(sigma_us_post, sigma_ch_post)

# Function to plot histograms
plot_boot_hist <- function(result, title, xlab_text) {
  hist(result$diffs, breaks = 30, main = title,
       xlab = xlab_text, col = "lightblue", border = "white")
  abline(v = result$mean_diff, col = "red", lwd = 2)
  text(result$mean_diff,
       max(hist(result$diffs, plot = FALSE)$counts) * 0.9,
       labels = sprintf("Observed diff = %.5f\np = %.3f",
                        result$mean_diff, result$pval),
       pos = 4, col = "red")
}

```


# Plots

```{r}
par(mfrow = c(2,2))
plot_boot_hist(boot_us_prepost, "US volatility: pre vs post COVID",
               "Mean diff (pre - post)")
plot_boot_hist(boot_ch_prepost, "China volatility: pre vs post COVID",
               "Mean diff (pre - post)")
plot_boot_hist(boot_us_ch_pre,  "US vs China: pre-COVID",
               "Mean diff (US - China)")
plot_boot_hist(boot_us_ch_post, "US vs China: post-COVID",
               "Mean diff (US - China)")
par(mfrow = c(1,1))


# Attach sigma to their Date vectors
# Make sure lengths match
stopifnot(length(sigma_us_pre)  == nrow(data_pre))
stopifnot(length(sigma_us_post) == nrow(data_post))
stopifnot(length(sigma_ch_pre)  == nrow(data_pre))
stopifnot(length(sigma_ch_post) == nrow(data_post))

df_pre  <- data.frame(Date = data_pre$Date,
                      sigma_us = sigma_us_pre,
                      sigma_china = sigma_ch_pre,
                      period = "pre")
df_post <- data.frame(Date = data_post$Date,
                      sigma_us = sigma_us_post,
                      sigma_china = sigma_ch_post,
                      period = "post")

# Combine
df_all <- rbind(df_pre, df_post)

# Aggregate duplicates by Date (take mean of values on same Date), this ensures only one value per calendar date
df_agg <- aggregate(cbind(sigma_us, sigma_china) ~ Date,
                    data = df_all,
                    FUN = mean, na.rm = TRUE)

# Add period labeling
# If the same calendar Date exists in pre & post, aggregated mean will merge them
df_agg <- df_agg[order(df_agg$Date), ]

# Define cutoff
cutoff <- as.Date("2020-01-01")
p2 <- ggplot(df_agg, aes(x = Date)) +
  geom_rect(aes(xmin = min(Date), xmax = cutoff - 1, ymin = -Inf, ymax = Inf),
            fill = "lightblue", alpha = 0.08) +
  geom_rect(aes(xmin = cutoff, xmax = max(Date), ymin = -Inf, ymax = Inf),
            fill = "pink", alpha = 0.04) +
  geom_line(aes(y = sigma_us, colour = "US"), size = 0.8) +
  geom_line(aes(y = sigma_china, colour = "China"), size = 0.8) +
  geom_vline(xintercept = as.numeric(cutoff), linetype = "dashed") +
  annotate("text", x = cutoff - 400, y = max(df_agg$sigma_us, na.rm=TRUE)*0.9, label = "Pre-COVID", size=3) +
  annotate("text", x = cutoff + 400, y = max(df_agg$sigma_us, na.rm=TRUE)*0.9, label = "Post-COVID", size=3) +
  labs(title = "Estimated conditional variance (sigma^2) with Pre/Post shading",
       y = expression(hat(sigma)^2),
       colour = "Series") +
  theme_minimal()
print(p2)

# Summary table
results_table <- data.frame(
  Comparison = c("US pre vs post",
                 "China pre vs post",
                 "US vs China (pre)",
                 "US vs China (post)"),
  Mean_Diff = c(boot_us_prepost$mean_diff,
                boot_ch_prepost$mean_diff,
                boot_us_ch_pre$mean_diff,
                boot_us_ch_post$mean_diff),
  P_Value = c(boot_us_prepost$pval,
              boot_ch_prepost$pval,
              boot_us_ch_pre$pval,
              boot_us_ch_post$pval)
)

print(results_table)
```
```{r}
params_us_pre  <- res_us_pre$par
params_us_post <- res_us_post$par
params_ch_pre  <- res_ch_pre$par
params_ch_post <- res_ch_post$par

# Function to compute persistence = alpha + beta
persistence <- function(par) par["alpha"] + par["beta"]

garch_table <- data.frame(
  Market = c("US", "US", "China", "China"),
  Period = c("Pre-COVID", "Post-COVID", "Pre-COVID", "Post-COVID"),
  Omega  = c(params_us_pre["omega"],
             params_us_post["omega"],
             params_ch_pre["omega"],
             params_ch_post["omega"]),
  Alpha  = c(params_us_pre["alpha"],
             params_us_post["alpha"],
             params_ch_pre["alpha"],
             params_ch_post["alpha"]),
  Beta   = c(params_us_pre["beta"],
             params_us_post["beta"],
             params_ch_pre["beta"],
             params_ch_post["beta"]),
  Persistence = c(persistence(params_us_pre),
                  persistence(params_us_post),
                  persistence(params_ch_pre),
                  persistence(params_ch_post))
)

kable(garch_table, digits = 6,
      caption = "GARCH(1,1) Parameter Comparison: Pre vs Post COVID") %>%
  kable_styling(full_width = FALSE, position = "center")
```



